---
title: "Recipes for multilevel imputation"
author: "Stef van Buuren (Utrecht University)"
date: "April 9, 2019"
output:
  beamer_presentation:
    colortheme: beaver
    includes:
      in_header: footline.tex
    theme: default
    toc: no
    keep_tex: true
  ioslides_presentation:
    fig_height: 1
    fig_width: 1
  slidy_presentation: default
classoption: aspectratio=43
---

```{r setup, include=FALSE, results = 'hide'}
library("knitr")
library("MASS", warn.conflicts = FALSE)
library("dplyr", warn.conflicts = FALSE)
suppressMessages(library("mice", warn.conflicts = FALSE))
suppressMessages(library("miceadds", warn.conflicts = FALSE))
suppressMessages(library("mitml", warn.conflicts = FALSE))
knitr::opts_chunk$set(echo = FALSE)
options(knitr.kable.NA = '')
source("R/hooks.R")
source("R/mi.hist.R")
knitr::read_chunk('R/fimd.R')
```

## Main question

*Can we use `mice` for multilevel data, and if so, how?*

## Imputation by fully conditional specification

```{r, fig.align='center'}
data <- expand.grid(rec = 1:8, var = 1:3)
data$r <- c(rep(3, 2), rep(1, 6), rep(3, 3), rep(1, 5), rep(1, 3), rep(3, 5))
levelplot(r~var+rec, data=data,
            as.table=FALSE, aspect="iso",
            shrink=c(0.9),
            col.regions = c(mdc(1), mdc(2), "grey90"), 
            cuts = 2,
            colorkey=FALSE,
            scales=list(draw=FALSE),
            xlab="", ylab="",
            between = list(x=1,y=0),
            strip = FALSE)
```


## Imputation by fully conditional specification

```{r, fig.align='center'}
data <- expand.grid(rec = 1:8, var = 1:3)
data$r <- c(rep(4, 2), rep(1, 6), rep(4, 3), rep(1, 5), rep(1, 3), rep(4, 5))
levelplot(r~var+rec, data=data,
            as.table=FALSE, aspect="iso",
            shrink=c(0.9),
            col.regions = c(mdc(1), mdc(2), "grey90", hcl(0, 100, 40, 0.3)), 
            cuts = 2,
            colorkey=FALSE,
            scales=list(draw=FALSE),
            xlab="", ylab="",
            between = list(x=1,y=0),
            strip = FALSE)
```

## Imputation by fully conditional specification

```{r, fig.align='center'}
data <- expand.grid(rec = 1:8, var = 1:3)
data$r <- c(rep(2, 2), rep(1, 6), rep(4, 3), rep(1, 5), rep(1, 3), rep(4, 5))
levelplot(r~var+rec, data=data,
            as.table=FALSE, aspect="iso",
            shrink=c(0.9),
            col.regions = c(mdc(1), mdc(2), "grey90", hcl(0, 100, 40, 0.3)), 
            cuts = 2,
            colorkey=FALSE,
            scales=list(draw=FALSE),
            xlab="", ylab="",
            between = list(x=1,y=0),
            strip = FALSE)
```

## Imputation by fully conditional specification

```{r, fig.align='center'}
data <- expand.grid(rec = 1:8, var = 1:3)
data$r <- c(rep(2, 2), rep(1, 6), rep(2, 3), rep(1, 5), rep(1, 3), rep(4, 5))
levelplot(r~var+rec, data=data,
            as.table=FALSE, aspect="iso",
            shrink=c(0.9),
            col.regions = c(mdc(1), mdc(2), "grey90", hcl(0, 100, 40, 0.3)), 
            cuts = 2,
            colorkey=FALSE,
            scales=list(draw=FALSE),
            xlab="", ylab="",
            between = list(x=1,y=0),
            strip = FALSE)
```

## Imputation by fully conditional specification

```{r, fig.align='center'}
data <- expand.grid(rec = 1:8, var = 1:3)
data$r <- c(rep(2, 2), rep(1, 6), rep(2, 3), rep(1, 5), rep(1, 3), rep(2, 5))
levelplot(r~var+rec, data=data,
            as.table=FALSE, aspect="iso",
            shrink=c(0.9),
            col.regions = c(mdc(1), mdc(2)), 
            cuts = 1,
            colorkey=FALSE,
            scales=list(draw=FALSE),
            xlab="", ylab="",
            between = list(x=1,y=0),
            strip = FALSE)
```

## Imputation by fully conditional specification - next iteration

```{r, fig.align='center'}
data <- expand.grid(rec = 1:8, var = 1:3)
data$r <- c(rep(4, 2), rep(1, 6), rep(4, 3), rep(1, 5), rep(1, 3), rep(4, 5))
levelplot(r~var+rec, data=data,
            as.table=FALSE, aspect="iso",
            shrink=c(0.9),
            col.regions = c(mdc(1), mdc(2), "grey90", hcl(0, 100, 40, 0.3)), 
            cuts = 2,
            colorkey=FALSE,
            scales=list(draw=FALSE),
            xlab="", ylab="",
            between = list(x=1,y=0),
            strip = FALSE)
```

## Imputation by fully conditional specification - next iteration

```{r, fig.align='center'}
data <- expand.grid(rec = 1:8, var = 1:3)
data$r <- c(rep(2, 2), rep(1, 6), rep(4, 3), rep(1, 5), rep(1, 3), rep(4, 5))
levelplot(r~var+rec, data=data,
            as.table=FALSE, aspect="iso",
            shrink=c(0.9),
            col.regions = c(mdc(1), mdc(2), "grey90", hcl(0, 100, 40, 0.3)), 
            cuts = 2,
            colorkey=FALSE,
            scales=list(draw=FALSE),
            xlab="", ylab="",
            between = list(x=1,y=0),
            strip = FALSE)
```



## `brandsma` data

- Brandsma and Knuver, Int J Ed Res, 1989.
- Extensively discussed in Snijders and Bosker (2012), 2nd ed.
- 4106 pupils, 216 schools, about 4% missing values

## `brandsma` data subset

```{r cache=TRUE, echo=TRUE}
library(mice)
d <- brandsma[, c("sch", "lpo", "sex", "den")]
head(d, 2)
```

- `sch`: School number, cluster variable, $C = 216$;
- `lpo`: Language test post, outcome at pupil level;
- `sex`: Sex of pupil, predictor at pupil level (0-1);
- `den`: School denomination, predictor at school level (1-4).

## Model of scientific interest
 
Predict `lpo` from the

- level-1 predictor `sex`
- level-2 predictor `den`

## Level notation - Bryk and Raudenbush (1992)

\begin{align}
{{\texttt{lpo}}}_{ic} & = \beta_{0c} + \beta_{1c}{{\texttt{sex}}}_{ic} + \epsilon_{ic}\\
\beta_{0c}     & = \gamma_{00} + \gamma_{01}{{\texttt{den}}}_{c} + u_{0c}\\
\beta_{1c}     & = \gamma_{10}
\end{align}

- $\text{lpo}_{ic}$ is the test score of pupil $i$ in school $c$
- $\text{sex}_{ic}$ is the sex of pupil $i$ in school $c$
- $\text{den}_c$ is the religious denomination of school $c$
- $\beta_{0c}$ is a random intercept that varies by cluster
- $\beta_{1c}$ is a sex effect, assumed to be the same across schools.
- $\epsilon_{ic} \sim N(0, \sigma_\epsilon^2)$ is the within-cluster random residual at the pupil level

## Where are the missings?

In single level data, missingness may be in the outcome and/or in the predictors

With multilevel data, missingness may be in:

1. the outcome variable;

2. the level-1 predictors;

3. the level-2 predictors;

4. the class variable.

## Univariate missing, level-1 outcome

```{r, fig.align='center'}
data <- expand.grid(rec = 1:8, var = 1:3)
data$r <- c(c(2, 1, 1), c(1, 1), c(2, 2, 1), 
            rep(1, 16))
levelplot(r~var+rec, data=data,
            as.table=FALSE, aspect="iso",
            shrink=c(0.9),
            col.regions = c(mdc(1), mdc(2)), 
            cuts = 1,
            colorkey=FALSE,
            scales = list(y = list(at = 1:8, labels = c(rep(3,3), rep(2,2), rep(1, 3)), tck = 0, cex = 2),
                          x = list(at = 1:3, labels = c("lpo", "sex", "den"), alt = 2, tck = 0, cex = 2)),
            xlab="", ylab="",
            between = list(x=10,y=0), strip = FALSE)
```

## Univariate missing, level-1 predictor, sporadically missing

```{r, fig.align='center'}
data <- expand.grid(rec = 1:8, var = 1:3)
data$r <- c(rep(1, 8), 
            c(1, 1, 2), c(1, 1), c(1, 2, 1), 
            rep(1, 8))
levelplot(r~var+rec, data=data,
            as.table=FALSE, aspect="iso",
            shrink=c(0.9),
            col.regions = c(mdc(1), mdc(2)), 
            cuts = 1,
            colorkey=FALSE,
            scales = list(y = list(at = 1:8, labels = c(rep(3,3), rep(2,2), rep(1, 3)), tck = 0, cex = 2),
                          x = list(at = 1:3, labels = c("lpo", "sex", "den"), alt = 2, tck = 0, cex = 2)),
            xlab="", ylab="",
            between = list(x=10,y=0), strip = FALSE)
```

## Univariate missing, level-1 predictor, systematically missing

```{r, fig.align='center'}
data <- expand.grid(rec = 1:8, var = 1:3)
data$r <- c(rep(1, 8), 
            c(1, 1, 2), c(1, 1), c(2, 2, 2), 
            rep(1, 8))
levelplot(r~var+rec, data=data,
            as.table=FALSE, aspect="iso",
            shrink=c(0.9),
            col.regions = c(mdc(1), mdc(2)), 
            cuts = 1,
            colorkey=FALSE,
            scales = list(y = list(at = 1:8, labels = c(rep(3,3), rep(2,2), rep(1, 3)), tck = 0, cex = 2),
                          x = list(at = 1:3, labels = c("lpo", "sex", "den"), alt = 2, tck = 0, cex = 2)),
            xlab="", ylab="",
            between = list(x=10,y=0), strip = FALSE)
```


## Univariate missing, level-2 predictor

```{r, fig.align='center'}
data <- expand.grid(rec = 1:8, var = 1:3)
data$r <- c(rep(1, 16), 
            c(1, 1, 1, 2, 2, 2, 2, 2))
levelplot(r~var+rec, data=data,
            as.table=FALSE, aspect="iso",
            shrink=c(0.9),
            col.regions = c(mdc(1), mdc(2)), 
            cuts = 1,
            colorkey=FALSE,
            scales = list(y = list(at = 1:8, labels = c(rep(3,3), rep(2,2), rep(1, 3)), tck = 0, cex = 2),
                          x = list(at = 1:3, labels = c("lpo", "sex", "den"), alt = 2, tck = 0), cex = 2),
            xlab="", ylab="",
            between = list(x=10,y=0), strip = FALSE)
```


## Multivariate missing

```{r, fig.align='center'}
data <- expand.grid(rec = 1:8, var = 1:3)
data$r <- c(c(2, 1, 1), c(1, 1), c(2, 2, 1),
            c(1, 1, 2), c(1, 1), c(1, 2, 1), 
            c(1, 1, 1, 2, 2, 2, 2, 2))
levelplot(r~var+rec, data=data,
            as.table=FALSE, aspect="iso",
            shrink=c(0.9),
            col.regions = c(mdc(1), mdc(2)), 
            cuts = 1,
            colorkey=FALSE,
            scales = list(y = list(at = 1:8, labels = c(rep(3,3), rep(2,2), rep(1, 3)), tck = 0, cex = 2),
                          x = list(at = 1:3, labels = c("lpo", "sex", "den"), alt = 2, tck = 0), cex = 2),
            xlab="", ylab="",
            between = list(x=10,y=0), strip = FALSE)
```


## Fully conditional specification for multilevel data

\begin{align}
\dot{{\texttt{lpo}}}_{ic} & \sim N(\beta_0 + \beta_1 {{\texttt{den}}}_{c} + \beta_2 {{\texttt{sex}}}_{ic} + u_{0c}, \sigma_\epsilon^2)\\
\dot{{\texttt{sex}}}_{ic} & \sim N(\beta_0 + \beta_1 {{\texttt{den}}}_{c} + \beta_2 {{\texttt{lpo}}}_{ic} + u_{0c}, \sigma_\epsilon^2)
\end{align}


## Theoretical problem with FCS

Conditional expectation of $\texttt{sex}_{ic}$ in a random effects model depends on

- $\texttt{lpo}_{ic}$,
- $\overline{\texttt{lpo}}_{i}$, the mean of cluster $i$, and
- $n_i$, the size of cluster $i$.

Resche-Rigon & White (2018) suggest the imputation model 

- should incorporate the cluster means of level-1 predictors 
- be heteroscedastic if cluster sizes vary

## General imputation/modeling sequence - START SIMPLE

1. Pick a simple complete-data model
2. Create imputations using an imputation template 
3. Check the imputes (convergence/plausibility) 
4. Estimate parameters
7. Make complete-data model more realistic, go to 1.

See <https://stefvanbuuren.name/fimd/sec-mlguidelines.html>

## Seven imputation templates, increasing complexity

1. *Intercept-only model, missing outcomes*
2. *Random intercepts, missing level-1 predictor*
3. Random intercepts, contextual model
4. *Random intercepts, missing level-2 predictor*
5. Random intercepts, interactions
6. Random slopes, missing outcomes and predictors
7. Random slopes, interactions

## 1 Intercept-only model, missing outcomes (model)

\begin{align}
{{\texttt{lpo}}}_{ic} & = \beta_{0c} + \epsilon_{ic}\\
\beta_{0c}     & = \gamma_{00} + u_{0c}
\end{align}

## 1 Intercept-only model, missing outcomes (imputation)

```{r mla.empty10, echo = TRUE}
```

## 1 Intercept-only model, missing outcomes (analysis)

```{r mla.empty5, echo = TRUE}
```

## 1 Intercept-only model, missing outcomes (variances)

```{r mla.empty6, echo = TRUE}
```

## 2 Random intercepts, missing level-1 (model)

\begin{align}
{{\texttt{lpo}}}_{ic} & = \beta_{0c} + \beta_{1c}{{\texttt{iqv}}}_{ic} + \epsilon_{ic}\\
\beta_{0c}     & = \gamma_{00} + u_{0c}\\
\beta_{1c}     & = \gamma_{10}
\end{align}

- Missing values in both `lpo` and `iqv`

## 2 Random intercepts, missing level-1 (imputation)

```{r mla.ri1, echo = TRUE, warning=FALSE}
```

- Impute `lpo` from `iqv` *and* the cluster means of `iqv`
- Impute `iqv` from `lpo` *and* the cluster means of `lpo`
- Alternative: Use `mitml::panImpute()` or `mitml::jomoImpute()`

## 2 Random intercepts, missing level-1 (predictorMatrix)

```{r echo = TRUE}
pred
```

## 2 Random intercepts, missing level-1 (analysis)

```{r mla.ri4, echo = TRUE}
```

## 4 Random intercepts, missing level-2 predictor (model)

\begin{align}
{{\texttt{lpo}}}_{ic} & = \beta_{0c} + \beta_{1c}{{\texttt{iqv}}}_{ic} + \epsilon_{ic}\\
\beta_{0c}     & = \gamma_{00} + \gamma_{01}{{\texttt{den}}}_{c} + u_{0c}\\
\beta_{1c}     & = \gamma_{10}
\end{align}

- Missing values in `lpo`, `iqv` and `den`
- For `den` the imputation model uses school level aggregates

## 4 Random intercepts, missing level-2 (imputation)

```{r mla.ril2p1, echo = TRUE, warning = FALSE}
```

## 4 Random intercepts, missing level-2 (predictorMatrix)

```{r echo = TRUE}
pred
```

## 4 Random intercepts, missing level-2 (density)

```{r mladenspmm, echo = FALSE, solo = TRUE}
p.pmm <- densityplot(imp, ~ lpo + den, ylab = "2l.pmm", ylim = list(c(0, 0.05), c(0, 1.5)))
p.pmm2 <- update(p.pmm, xlim = list(c(16, 64), c(0, 5)))
gridExtra::grid.arrange(p.pmm2)
```

## 4 Random intercepts, missing level-2 (analysis)

```{r mla.ril2p6}
```

## Classic recipe for single-level data: Which predictors?

1. Include all variables that appear in the complete-data model
2. Include variables related to the nonresponse
3. Include variables that explain a considerable amount of variance
4. Remove from variables selected in steps 2 and 3 those variables
   that have too many missing values within the subgroup of
   incomplete cases

*Does this recipe also apply to multilevel data?*


## Recipe: Missing level-1

     Recipe for a level-1 target
---- ---------------------------------------------------------------
1.   Define the most general analytic model 
2.   Select a `2l` method that imputes close to the data
3.   Include all level-1 variables
4.   Include the disaggregated cluster means of level-1 variables
5.   Include all level-1 interactions implied by analytic model
6.   Include all level-2 predictors
7.   Include all level-2 interactions implied by analytic model
8.   Include all cross-level interactions implied by analytic model
9.   Include predictors related to the missingness and the target
10.  Exclude any terms involving the target

## Recipe: Missing level-2

     Recipe for a level-2 target
---- ---------------------------------------------------------------
1.   Define the most general analytic model
2.   Select a `2lonly` method that imputes close to the data
3.   Include the cluster means of all level-1 variables
4.   Include the cluster means of all level-1 interactions
5.   Include all level-2 predictors
6.   Include all interactions of level-2 variables
7.   Include predictors related to the missingness and target
8.   Exclude any terms involving the target

## Conclusion

*Can we use `mice` for multilevel data, and if so, how?*

- Hot spot of current research
- Multilevel imputation: more complex, but doable
- Start simple, take small steps
- Build upon templates and modeling recipes
- Study <https://stefvanbuuren.name/fimd/sec-mlguidelines.html>
- Gain confidence at each step
- Start playing around...
